---
title: "Discovering Available Layers"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Discovering Available Layers}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

This vignette shows how to inspect the Wildfire Resilience Index (WRI)
catalog before downloading data. Users start by listing available layers,
then narrow the catalog to the layer ID they need. Catalog browsing reads
the local STAC index bundled with the package, so this step does not
require a large data download.

## Setup

Load firex. No other packages are needed for catalog browsing.

```{r}
library(firex)
```

## Browsing the Full Catalog

`wri_overview()` reads the local STAC index and prints a structured
summary of all 82 available layers. Call it with no arguments to see all
domains, dimensions, and data types at a glance:

```{r}
wri_overview()
```

```text
WRI DATA SUMMARY
-------------------
Read at:      2026-05-11 16:07:50

Layer ID Naming Convention: wri_domain + wri_dimension + data_type

Collections (1): wri_ignitR

Unique wri_domain values (11): air_quality, communities, iconic_places,
iconic_species, infrastructure, livelihoods, natural_habitats,
sense_of_place, species, unknown, water

Unique data_type values (3): aggregate, final_score, indicator

Unique wri_dimension values (5): domain_score, recovery, resilience,
resistance, status

Total layers available: 82
```

The object summarizes the 82 available layers grouped by domain,
dimension, and data type.

## Getting the Catalog as a Data Frame

`wri_overview_df()` returns the same information as a flat data frame,
one row per layer. This is useful for filtering with standard R tools:

```{r}
df <- wri_overview_df()
nrow(df)
#> [1] 82

# Filter to water-domain aggregate layers
subset(
  df,
  wri_domain == "water" & data_type == "aggregate",
  select = c(id, wri_dimension, data_type)
)
#>                    id wri_dimension data_type
#> 75 water_domain_score  domain_score aggregate
#> 76   water_resilience    resilience aggregate
#> 77   water_resistance    resistance aggregate
#> 80       water_status        status aggregate
```

The ten named WRI domains are `air_quality`, `communities`,
`iconic_places`, `iconic_species`, `infrastructure`, `livelihoods`,
`natural_habitats`, `sense_of_place`, `species`, and `water`. The
`unknown` domain is used for the overall `WRI_score` layer. The five
WRI dimension values are `domain_score`, `recovery`, `resilience`,
`resistance`, and `status`.

## Inspecting a Single Layer

Once you have identified a layer ID, `layer_info()` retrieves its full
metadata record, including the hosted COG URL, geographic extent, CRS,
and classification properties. Transposing with `t()` makes it easier
to read:

```{r}
t(layer_info("WRI_score"))
#>            WRI_score
#> id         WRI_score
#> collection wri_ignitR
#> asset_href https://knb.ecoinformatics.org/.../WRI_score.tif
#> proj_code  EPSG:5070
#> data_type  final_score
#> wri_domain unknown
#> is_hosted  TRUE
#> xmin       -146.2082
#> ymin        19.1074
#> xmax        173.7109
#> ymax         54.8056
```

The `asset_href` field is the direct URL to the hosted COG file on KNB.
`get_layer()` uses this URL to open a virtual connection via GDAL; you
do not need to use it directly. The `xmin`, `ymin`, `xmax`, and `ymax`
fields give the layer's geographic extent in WGS84 (EPSG:4326).

## Advanced: Querying STAC Items Directly

For filtering on combinations of properties not easily expressed as data
frame queries, `query_stac_flexible()` filters the underlying STAC item
list by any property value:

```{r}
catalog <- wri_overview()
items <- lapply(catalog$data$items, function(x) x$item)

# All resistance indicators in the water domain
water_resist <- query_stac_flexible(
  items,
  wri_domain = "water",
  wri_dimension = "resistance",
  data_type = "indicator"
)

vapply(water_resist, function(x) x$id, character(1))
#> [1] "water_resistance_drought_plans"
#> [2] "water_resistance_water_treatment"
```

Once users identify a layer ID, they can pass that ID to `get_layer()`
or use an existing spatial object as the area of interest.
