---
title: "firex Package Demo"
format: html
---

# Getting Started

This demo shows how to explore, query, and retrieve WRI data using the `firex` package.

```{r setup, message=FALSE}
# Load the package (development mode)
# If you've installed the package, use: library(firex)
# For development from the package root, use:
devtools::load_all()

# Load terra for plotting
library(terra)
```

## Step 1: Explore What's Available

`wri_overview()` reads the local STAC catalog and prints a summary of the
available domains, dimensions, and data types.

```{r overview}
catalog <- wri_overview()
catalog
```

`wri_overview_df()` returns a flat data frame that is convenient for quick
filtering in base R.

```{r overview-df}
catalog_df <- wri_overview_df()
head(catalog_df[, c("id", "wri_domain", "wri_dimension", "data_type", "is_hosted")])
```

You can still inspect the categories directly from the flattened catalog:

```{r explore-categories}
sort(unique(catalog_df$wri_domain))
sort(unique(catalog_df$wri_dimension))
sort(unique(catalog_df$data_type))
```

## Step 2: Filter and Inspect Layers

Filter the catalog data frame to find layers you want to work with:

```{r filter-df}
# All water layers
subset(
  catalog_df,
  wri_domain == "water",
  select = c(id, wri_dimension, data_type)
)

# Indicator layers for air quality
subset(
  catalog_df,
  wri_domain == "air_quality" & data_type == "indicator",
  select = c(id, wri_dimension, data_type)
)

# Search layer ids by name pattern
catalog_df[grepl("surface_water|asthma", catalog_df$id), c("id", "wri_domain", "data_type")]
```

Use `layer_info()` when you want the full metadata for a single layer:

```{r layer-info}
layer_info("water_status_surface_water_quantity")
```

Use `query_stac_flexible()` when you want to filter the underlying STAC items by
property values:

```{r query-stac}
water_indicator_items <- query_stac_flexible(
  catalog$data$items,
  wri_domain = "water",
  data_type = "indicator"
)

vapply(water_indicator_items, function(x) x$id, character(1))
```

## Step 3: Retrieve and Plot Data

### Example 1: Get the Full WRI Score Layer

```{r get-layer-full}
# Returns a SpatRaster pointing at the hosted COG
t_full <- system.time({
  wri_full <- get_layer("WRI_score")
})
t_full
```

```{r plot-layer-full}
# Plotting the full layer can take a while, so it is left commented out.
#plot(wri_full,
#     main = "WRI Score - Full Extent",
#     col = hcl.colors(100, "YlOrRd", rev = TRUE))
```

### Example 2: Get WRI Score for Northern California

```{r get-layer-bbox}
# Bounding boxes use c(xmin, ymin, xmax, ymax) in the supplied CRS.
bbox_norcal <- c(-122, 37, -121, 38)

t_norcal <- system.time({
  wri_norcal <- get_layer("WRI_score", bbox = bbox_norcal)
})
t_norcal
```

```{r plot-layer-bbox}
plot(wri_norcal,
     main = "WRI Score - Northern California",
     col = hcl.colors(100, "YlOrRd", rev = TRUE))
```

### Example 3: Retrieve a Specific Indicator Layer

```{r get-layer-indicator}
indicator_id <- "water_status_surface_water_quantity"

t_indicator <- system.time({
  water_quantity_norcal <- get_layer(indicator_id, bbox = bbox_norcal)
})
t_indicator
```

```{r plot-layer-indicator}
plot(water_quantity_norcal,
     main = "Surface Water Quantity - Northern California",
     col = hcl.colors(100, "YlOrRd", rev = TRUE))
```

## Summary

1. Explore the catalog with `wri_overview()` and `wri_overview_df()`.
2. Filter layers with base R, `layer_info()`, and `query_stac_flexible()`.
3. Retrieve hosted rasters with `get_layer(id, bbox = ...)`.

