---
title: "firex Package Demo"
format: html
---

# firex Package Demo

This document demonstrates the core capabilities of the `firex` package:
browsing available WRI Aqueduct layers, retrieving raster data, and the
range of spatial input types that `get_layer()` accepts. It also shows how
the package handles common CRS problems with clear error messages.

Chunks can be run individually. The setup chunk below must be run first to
load packages and prepare the demo data objects.

## Data Sources

The spatial examples below use Eaton Fire perimeter data from LA County Open Data.

| File | Format | Source |
|------|--------|--------|
| `Eaton_Perimeter_20250121.shp` | Shapefile | [LA County Open Data - Eaton Fire Perimeter](https://data.lacounty.gov/datasets/lacounty::eaton-fire-perimeter/about) |
| `Eaton_Fire_Perimeter.geojson` | GeoJSON | [LA County Open Data - Eaton Fire Perimeter](https://data.lacounty.gov/datasets/lacounty::eaton-fire-perimeter/about) |

Both files are bundled in `inst/demos/data/` and loaded via `system.file()`.

---

## Setup

```{r setup, message=FALSE}
# Install from GitHub (run once, then comment out):
# devtools::install_github("ignitR-package/firex")

# Or load from source during development:
# devtools::load_all()

library(firex)
library(terra)
library(sf)

# Paths to bundled demo data
data_dir   <- system.file("demos/data", package = "firex")
shp_path   <- file.path(data_dir, "Eaton_Perimeter_20250121.shp")
gjson_path <- file.path(data_dir, "Eaton_Fire_Perimeter.geojson")

# Load the shapefile as different object types for the examples below
eaton_sf   <- sf::st_read(shp_path, quiet = TRUE)   # sf object
eaton_vect <- terra::vect(shp_path)                  # SpatVector
eaton_crs  <- terra::crs(eaton_vect)                 # save CRS before ext() drops it
eaton_ext  <- terra::ext(eaton_vect)                 # SpatExtent (no CRS attached)
```

---

## Part 1: Browse the Catalog

Use `wri_overview()` to see what layers are available, and `wri_overview_df()`
to get that information as a data frame you can filter.

```{r overview}
catalog    <- wri_overview()
catalog_df <- wri_overview_df()

# Preview available layers
head(catalog_df[, c("id", "wri_domain", "wri_dimension", "data_type", "is_hosted")])
```

```{r explore-categories}
# See all available domains, dimensions, and data types
sort(unique(catalog_df$wri_domain))
sort(unique(catalog_df$wri_dimension))
sort(unique(catalog_df$data_type))
```

Filter the data frame to find layers of interest:

```{r filter-df}
# All water layers
subset(catalog_df, wri_domain == "water",
       select = c(id, wri_dimension, data_type))

# Air quality indicator layers only
subset(catalog_df, wri_domain == "air_quality" & data_type == "indicator",
       select = c(id, wri_dimension, data_type))

# Search layer IDs by keyword
catalog_df[grepl("surface_water|asthma", catalog_df$id),
           c("id", "wri_domain", "data_type")]
```

Use `layer_info()` to inspect the full metadata for a single layer:

```{r layer-info}
layer_info("water_status_surface_water_quantity")
```

Use `query_stac_flexible()` to filter the underlying STAC items directly by
any property value:

```{r query-stac}
stac_items <- lapply(catalog$data$items, function(x) x$item)

water_items <- query_stac_flexible(stac_items,
                                   wri_domain = "water",
                                   data_type  = "indicator")

vapply(water_items, function(x) x$id, character(1))
```

---

## Part 2: Retrieve Data with `get_layer()`

`get_layer(id, aoi, aoi_crs)` streams a hosted raster layer and optionally
crops it to a spatial area of interest (`aoi`). The `aoi_crs` argument tells
the function what coordinate reference system your `aoi` is in -- more on
that below.

The function returns a `SpatRaster` (from the `terra` package).

### No AOI - full layer

Omit `aoi` to retrieve the full global extent of a layer. This can be slow
to plot, so the plot call is commented out.

```{r get-full}
wri_full <- get_layer("WRI_score")

# plot(wri_full,
#      main = "WRI Score - Full Extent",
#      col  = hcl.colors(100, "YlOrRd", rev = TRUE))
```

---

### AOI type 1: numeric bounding box

Pass a vector `c(xmin, ymin, xmax, ymax)` as the `aoi`. You must also supply
`aoi_crs` to tell the function what coordinate system those numbers are in.

```{r get-bbox}
wri_norcal <- get_layer("WRI_score",
                         aoi     = c(-122, 37, -121, 38),
                         aoi_crs = "EPSG:4326")

plot(wri_norcal,
     main = "WRI Score - Northern California",
     col  = hcl.colors(100, "YlOrRd", rev = TRUE))
```

---

### AOI type 2: shapefile path

Pass a file path to a `.shp` file. The CRS is read from the `.prj` sidecar
file, so `aoi_crs` is not needed.

```{r get-shp}
# Source: LA County Open Data - Eaton Fire Perimeter
wri_shp <- get_layer("WRI_score", aoi = shp_path)

plot(wri_shp,
     main = "WRI Score - Eaton Fire Area (shapefile)",
     col  = hcl.colors(100, "YlOrRd", rev = TRUE))
```

---

### AOI type 3: GeoJSON path

Pass a file path to a `.geojson` file. CRS is embedded in the file format.

```{r get-geojson}
# Source: LA County Open Data - Eaton Fire Perimeter
wri_gjson <- get_layer("WRI_score", aoi = gjson_path)

plot(wri_gjson,
     main = "WRI Score - Eaton Fire Area (GeoJSON)",
     col  = hcl.colors(100, "YlOrRd", rev = TRUE))
```

---

### AOI type 4: sf object

Pass an `sf` object directly. The CRS is embedded in the object.

```{r get-sf}
# eaton_sf was loaded in setup via sf::st_read()
wri_sf <- get_layer("WRI_score", aoi = eaton_sf)

plot(wri_sf,
     main = "WRI Score - Eaton Fire Area (sf object)",
     col  = hcl.colors(100, "YlOrRd", rev = TRUE))
```

---

### AOI type 5: SpatVector

Pass a `terra` SpatVector. The CRS is embedded in the object. This example
also shows using a different layer ID.

```{r get-spatvector}
# eaton_vect was loaded in setup via terra::vect()
water_qty <- get_layer("water_status_surface_water_quantity",
                        aoi = eaton_vect)

plot(water_qty,
     main = "Surface Water Quantity - Eaton Fire Area (SpatVector)",
     col  = hcl.colors(100, "YlOrRd", rev = TRUE))
```

---

### AOI type 6: SpatExtent

`terra::ext()` returns a `SpatExtent` -- a plain bounding box with no CRS
attached. You must supply `aoi_crs` explicitly.

```{r get-spatextent}
# eaton_ext was created in setup via terra::ext(eaton_vect)
# eaton_crs was saved before ext() stripped it
wri_ext <- get_layer("WRI_score",
                      aoi     = eaton_ext,
                      aoi_crs = eaton_crs)

plot(wri_ext,
     main = "WRI Score - Eaton Fire Area (SpatExtent)",
     col  = hcl.colors(100, "YlOrRd", rev = TRUE))
```

---

## Part 3: CRS Handling and Error Messages

`get_layer()` validates the AOI before making any network request. The
examples below show what happens when the CRS information is missing,
conflicting, or incorrect. All calls are wrapped in `tryCatch` so they print
the error message without stopping the script.

### Missing CRS: numeric bbox without `aoi_crs`

When you pass a numeric vector as the AOI, `aoi_crs` is required. Without it
the function has no way to know what coordinate system the numbers refer to.

```{r error-no-crs}
tryCatch(
  get_layer("WRI_score", aoi = c(-118.2, 34.1, -117.9, 34.3)),
  error = function(e) message("Error: ", conditionMessage(e))
)
```

---

### Missing CRS: SpatExtent without `aoi_crs`

A `SpatExtent` also carries no CRS (terra strips it). The same rule applies.

```{r error-extent-no-crs}
tryCatch(
  get_layer("WRI_score", aoi = eaton_ext),
  error = function(e) message("Error: ", conditionMessage(e))
)
```

---

### Conflicting CRS: `aoi_crs` supplied when the object already has one

When you pass an `sf` object or `SpatVector`, the CRS is already embedded.
Supplying `aoi_crs` as well is ambiguous, so the function rejects it.

```{r error-crs-conflict}
tryCatch(
  get_layer("WRI_score", aoi = eaton_sf, aoi_crs = "EPSG:4326"),
  error = function(e) message("Error: ", conditionMessage(e))
)
```

---

### Invalid layer ID

If the layer ID is not in the catalog, the function stops with a clear
message rather than failing silently.

```{r error-bad-id}
tryCatch(
  get_layer("this_layer_does_not_exist"),
  error = function(e) message("Error: ", conditionMessage(e))
)
```

---

### Warning: AOI partially overlaps the layer extent

If the AOI extends beyond the edge of the layer, `get_layer()` issues a
warning naming which sides overflow, then crops to the available area and
returns the result normally.

```{r warn-partial-overlap, warning=TRUE}
# This bbox extends well beyond the contiguous US
wri_overflow <- get_layer("WRI_score",
                           aoi     = c(-130, 30, -55, 55),
                           aoi_crs = "EPSG:4326")
```

---

### Error: AOI does not overlap the layer at all

If there is no overlap, the function stops.

```{r error-no-overlap}
tryCatch(
  get_layer("WRI_score",
             aoi     = c(-80, -90, -60, -70),
             aoi_crs = "EPSG:4326"),
  error = function(e) message("Error: ", conditionMessage(e))
)
```

---

## Summary

### Functions

| Function | Purpose |
|----------|---------|
| `wri_overview()` | Print a summary of available layers |
| `wri_overview_df()` | Return available layers as a data frame |
| `layer_info(id)` | Full metadata for one layer |
| `query_stac_flexible(items, ...)` | Filter STAC items by property |
| `get_layer(id, aoi, aoi_crs)` | Stream a hosted raster, cropped to AOI |

### AOI input types for `get_layer()`

| AOI type | `aoi_crs` required? |
|----------|---------------------|
| `NULL` (returns full layer) | No |
| Numeric `c(xmin, ymin, xmax, ymax)` | Yes, always |
| File path to `.shp` | No (read from `.prj` sidecar) |
| File path to `.geojson` | No (embedded in file) |
| `sf` or `sfc` object | No (embedded in object) |
| `terra::SpatVector` | No (embedded in object) |
| `terra::SpatExtent` | Yes, always (terra strips the CRS) |
