---
title: "firex Package Demo"
format: html
---

# firex Package Demo

This document demonstrates the core capabilities of the `firex` package:
browsing available WRI Aqueduct layers, retrieving raster data, and the range of spatial input types that `get_layer()` accepts. It also shows how the package handles common CRS problems with clear error messages.

Chunks can be run individually. The setup chunk below must be run first to load packages and prepare the demo data objects.

## Data Sources

The spatial examples below use Eaton Fire perimeter data from LA County Open Data.

| File | Format | Source |
|------|--------|--------|
| `Eaton_Perimeter_20250121.shp` | Shapefile | [LA County Open Data - Eaton Fire Perimeter](https://data.lacounty.gov/datasets/lacounty::eaton-fire-perimeter/about) |
| `Eaton_Fire_Perimeter.geojson` | GeoJSON | [LA County Open Data - Eaton Fire Perimeter](https://data.lacounty.gov/datasets/lacounty::eaton-fire-perimeter/about) |

Both files are bundled in `inst/demos/data/` and loaded via `system.file()`.

---

## Setup

```{r setup, message=FALSE}
# Install from GitHub (run once, then comment out):
# devtools::install_github("ignitR-package/firex")

# Or load from source during development:
devtools::load_all()

library(terra)
library(sf)

# Paths to bundled demo data
data_dir   <- system.file("demos/data", package = "firex")
shp_path   <- file.path(data_dir, "Eaton_Perimeter_20250121.shp")
gjson_path <- file.path(data_dir, "Eaton_Fire_Perimeter.geojson")

# Load the shapefile as different object types for the examples below
eaton_sf   <- sf::st_read(shp_path, quiet = TRUE)   # sf object
eaton_vect <- terra::vect(shp_path)                  # SpatVector
eaton_crs  <- terra::crs(eaton_vect)                 # save CRS before ext() drops it
eaton_ext  <- terra::ext(eaton_vect)                 # SpatExtent (no CRS attached)
```

## Part 1: Browse the Catalog

Use `wri_overview()` to see what layers are available, and `wri_overview_df()` to get that information as a data frame you can filter.

```{r overview}
catalog    <- wri_overview()
catalog_df <- wri_overview_df()

# Preview available layers
head(catalog_df[, c("id", "wri_domain", "wri_dimension", "data_type", "is_hosted")])
```

```{r explore-categories}
# See all available domains, dimensions, and data types
cat("Domains: \n")
sort(unique(catalog_df$wri_domain))
cat("\nDimensions: \n")
sort(unique(catalog_df$wri_dimension))
cat("\nData Types: \n")
sort(unique(catalog_df$data_type))
```

Filter the data frame to find layers of interest:

```{r filter-df}
# All water layers
subset(catalog_df, wri_domain == "water",
       select = c(id, wri_dimension, data_type))

# Air quality indicator layers only
subset(catalog_df, wri_domain == "air_quality" & data_type == "indicator",
       select = c(id, wri_dimension, data_type))
```

Use `layer_info()` to inspect the full metadata for a single layer:

```{r layer-info}
t(layer_info("water_status_surface_water_quantity"))
```

Use `query_stac_flexible()` to filter the underlying STAC items directly by any property value:

```{r query-stac}
stac_items <- lapply(catalog$data$items, function(x) x$item)

water_items <- query_stac_flexible(stac_items,
                                   wri_domain = "water",
                                   data_type  = "indicator")

vapply(water_items, function(x) x$id, character(1))
```

---

## Part 2: Retrieve Data with `get_layer()`

`get_layer(id, aoi, aoi_crs)` streams a hosted raster layer and optionally crops it to a spatial area of interest (`aoi`). The `aoi_crs` argument tells the function what coordinate reference system your `aoi` is in (more on that below).

The function returns a `SpatRaster` (from the `terra` package).

### No AOI - full layer

Omit `aoi` to retrieve the full global extent of a layer. Plotting the full WRI_score layer takes ~30-60 seconds. The `get_layer()` call itself is usually fast here because `terra` opens the remote COG lazily and only reads header metadata until you actually crop, plot, or extract values.

```{r get-full}
wri_full <- get_layer("WRI_score")

dir.create("demo_output", showWarnings = FALSE)
png("demo_output/wri_score_full.png", width = 1200, height = 900, res = 150)
system.time({
  plot(wri_full,
       main = "WRI Score - Full Extent",
       col  = hcl.colors(100, "YlOrRd", rev = TRUE))
})['elapsed']
dev.off()
```

---

### AOI type 1: numeric bounding box

Pass a vector `c(xmin, ymin, xmax, ymax)` as the `aoi`. You must also supply `aoi_crs` to tell the function what coordinate system those numbers are in. The returned object keeps a lazy window on the remote COG, so the call itself is usually fast and the real read happens when you plot, extract values, or write the raster.

```{r get-bbox}
wri_norcal <- get_layer("WRI_score",
                         aoi     = c(-122, 37, -121, 38),
                         aoi_crs = "EPSG:4326")

system.time({
  plot(wri_norcal,
     main = "WRI Score - Northern California",
     col  = hcl.colors(100, "YlOrRd", rev = TRUE))
})['elapsed']

```

---

### AOI type 2: shapefile path

Pass a file path to a `.shp` file. The CRS is read from the `.prj` sidecar file, so `aoi_crs` is not needed.

```{r get-shp}
# Source: LA County Open Data - Eaton Fire Perimeter
wri_shp <- get_layer("WRI_score", aoi = shp_path)

system.time({
  plot(wri_shp,
     main = "WRI Score - Eaton Fire Area (shapefile)",
     col  = hcl.colors(100, "YlOrRd", rev = TRUE))
})['elapsed']
plot(terra::project(terra::vect(shp_path), terra::crs(wri_shp)),
     add = TRUE, border = "black")

```

---

### AOI type 3: GeoJSON path

Pass a file path to a `.geojson` file. CRS is embedded in the file format.

```{r get-geojson}
# Source: LA County Open Data - Eaton Fire Perimeter
wri_gjson <- get_layer("WRI_score", aoi = gjson_path)

system.time({})['elapsed']
plot(wri_gjson,
     main = "WRI Score - Eaton Fire Area (GeoJSON)",
     col  = hcl.colors(100, "YlOrRd", rev = TRUE))
plot(terra::project(terra::vect(gjson_path), terra::crs(wri_gjson)),
     add = TRUE, border = "black")
```

---

### AOI type 4: sf object

Pass an `sf` object directly. The CRS is embedded in the object.

```{r get-sf}
# eaton_sf was loaded in setup via sf::st_read()
wri_sf <- get_layer("WRI_score", aoi = eaton_sf)

plot(wri_sf,
     main = "WRI Score - Eaton Fire Area (sf object)",
     col  = hcl.colors(100, "YlOrRd", rev = TRUE))
plot(terra::project(terra::vect(eaton_sf), terra::crs(wri_sf)),
     add = TRUE, border = "black")
```

---

### AOI type 5: SpatVector

Pass a `terra` SpatVector. The CRS is embedded in the object. This example also shows using a different layer ID.

```{r get-spatvector}
# eaton_vect was loaded in setup via terra::vect()
water_qty <- get_layer("infrastructure_resistance",
                        aoi = eaton_vect)

plot(water_qty,
     main = "Infrastructure Resistance - Eaton Fire Area (SpatVector)",
     col  = hcl.colors(100, "YlOrRd", rev = TRUE))
plot(terra::project(eaton_vect, terra::crs(water_qty)),
     add = TRUE, border = "black")
```

---

### AOI type 6: SpatExtent

`terra::ext()` returns a `SpatExtent` -- a plain bounding box with no CRS attached. You must supply `aoi_crs` explicitly.

```{r get-spatextent}
# eaton_ext was created in setup via terra::ext(eaton_vect)
# eaton_crs was saved before ext() stripped it
wri_ext <- get_layer("WRI_score",
                      aoi     = eaton_ext,
                      aoi_crs = eaton_crs)

plot(wri_ext,
     main = "WRI Score - Eaton Fire Area (SpatExtent)",
     col  = hcl.colors(100, "YlOrRd", rev = TRUE))
plot(terra::project(terra::vect(eaton_ext, crs = eaton_crs), terra::crs(wri_ext)),
     add = TRUE, border = "black")
```

---

## Part 3: CRS Handling and Error Messages

`get_layer()` validates the AOI before making any network request. The
examples below show what happens when the CRS information is missing,
conflicting, or incorrect. All calls are wrapped in `tryCatch` so they print the error message without stopping the script.

### Missing CRS: numeric bbox without `aoi_crs`

When you pass a numeric vector as the AOI, `aoi_crs` is required. Without it the function has no way to know what coordinate system the numbers refer to.

```{r error-no-crs}
tryCatch(
  get_layer("WRI_score", aoi = c(-118.2, 34.1, -117.9, 34.3)),
  error = function(e) message("Error: ", conditionMessage(e))
)
```

---

### Missing CRS: SpatExtent without `aoi_crs`

A `SpatExtent` also carries no CRS (terra strips it). The same rule applies.

```{r error-extent-no-crs}
tryCatch(
  get_layer("WRI_score", aoi = eaton_ext),
  error = function(e) message("Error: ", conditionMessage(e))
)
```

---

### Conflicting CRS: `aoi_crs` supplied when the object already has one

When you pass an `sf` object or `SpatVector`, the CRS is already embedded. Supplying `aoi_crs` as well is ambiguous, so the function rejects it.

```{r error-crs-conflict}
tryCatch(
  get_layer("WRI_score", aoi = eaton_sf, aoi_crs = "EPSG:4326"),
  error = function(e) message("Error: ", conditionMessage(e))
)
```

---

### Invalid layer ID

If the layer ID is not in the catalog, the function stops with a clear
message rather than failing silently.

```{r error-bad-id}
tryCatch(
  get_layer("this_layer_does_not_exist"),
  error = function(e) message("Error: ", conditionMessage(e))
)
```

---

### Warning: AOI partially overlaps the layer extent

If the AOI extends beyond the edge of the layer, `get_layer()` issues a warning naming which sides overflow, then intersects the request with the available area and returns the result normally.

```{r warn-partial-overlap, warning=TRUE}
# This bbox covers most of the contiguous US, requires crop so does not load lazily, ~35 seconds again
system.time({
  wri_overflow <- get_layer("WRI_score",
                           aoi     = c(-130, 30, -55, 55),
                           aoi_crs = "EPSG:4326")
})['elapsed']

```

---

### Error: AOI does not overlap the layer at all

If there is no overlap, the function stops.

```{r error-no-overlap}
tryCatch(
  get_layer("WRI_score",
             aoi     = c(-80, -90, -60, -70),
             aoi_crs = "EPSG:4326"),
  error = function(e) message("Error: ", conditionMessage(e))
)
```

---

## Part 4: Compare All-Layer Extract Sizes

This section measures how large the full set of hosted WRI raster layers would be for several extraction regions. The first three regions use WGS84 bounding boxes. The full-layer case uses the same default behavior as `get_layer()` with no `aoi`: it records the hosted COG file size from the remote headers instead of rewriting every full raster locally.

The chunk is not evaluated automatically because it opens every hosted WRI layer for every region and writes many raster extracts. Run it after the setup chunk when you want to generate the CSV summaries and bar graph.

```{r wri-extract-size-comparison, eval=FALSE}
# Bounding boxes are c(xmin, ymin, xmax, ymax) in EPSG:4326.
# Adjust the Southern California bbox if your analysis uses a different boundary.
wri_size_regions <- list(
  "Santa Barbara County" = list(
    aoi = c(-120.7, 34.4, -119.4, 35.1),
    aoi_crs = "EPSG:4326"
  ),
  "Southern California" = list(
    aoi = c(-121.0, 32.4, -114.0, 36.0),
    aoi_crs = "EPSG:4326"
  ),
  "California" = list(
    aoi = c(-124.6, 32.4, -114.0, 42.1),
    aoi_crs = "EPSG:4326"
  ),
  "Full layer extent" = list(
    aoi = NULL,
    aoi_crs = NULL
  )
)

demo_dir <- if (dir.exists(file.path("inst", "demos"))) {
  file.path("inst", "demos")
} else {
  "."
}
demo_output_dir <- file.path(demo_dir, "demo_output")
extract_cache_dir <- file.path(demo_output_dir, "wri_extract_size_cache")
detail_csv <- file.path(demo_output_dir, "wri_extract_file_sizes.csv")
totals_csv <- file.path(demo_output_dir, "wri_extract_size_totals.csv")
plot_path <- file.path(demo_output_dir, "wri_extract_size_totals.png")
wri_score_plot_dir <- file.path(demo_output_dir, "wri_score_region_plots")

dir.create(extract_cache_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(wri_score_plot_dir, recursive = TRUE, showWarnings = FALSE)
Sys.setenv(
  GDAL_DISABLE_READDIR_ON_OPEN = "EMPTY_DIR",
  CPL_VSIL_CURL_ALLOWED_EXTENSIONS = ".tif"
)

safe_name <- function(x) {
  gsub("[^A-Za-z0-9_-]+", "_", x)
}

remote_file_size <- function(url) {
  headers <- tryCatch(
    curlGetHeaders(url, redirect = TRUE),
    error = function(e) character()
  )

  content_length <- grep(
    "^content-length:",
    headers,
    value = TRUE,
    ignore.case = TRUE
  )

  if (length(content_length) == 0L) {
    return(NA_real_)
  }

  as.numeric(sub("^[^:]+:\\s*", "", tail(content_length, 1L)))
}

measure_one_extract <- function(layer_id, href, region_name, region) {
  if (is.null(region$aoi)) {
    bytes <- remote_file_size(href)

    return(data.frame(
      region = region_name,
      id = layer_id,
      file_path = href,
      bytes = bytes,
      error = if (is.na(bytes)) "Could not read Content-Length header" else NA_character_,
      stringsAsFactors = FALSE
    ))
  }

  region_dir <- file.path(extract_cache_dir, safe_name(region_name))
  out_path <- file.path(region_dir, paste0(layer_id, ".tif"))
  dir.create(region_dir, recursive = TRUE, showWarnings = FALSE)

  if (!file.exists(out_path)) {
    raster <- get_layer(
      layer_id,
      aoi = region$aoi,
      aoi_crs = region$aoi_crs
    )

    terra::writeRaster(
      raster,
      out_path,
      overwrite = TRUE,
      gdal = c("COMPRESS=DEFLATE", "TILED=YES", "BIGTIFF=IF_SAFER")
    )
  }

  data.frame(
    region = region_name,
    id = layer_id,
    file_path = out_path,
    bytes = file.info(out_path)$size,
    error = NA_character_,
    stringsAsFactors = FALSE
  )
}

save_wri_score_plot <- function(region_name, region) {
  out_path <- file.path(
    wri_score_plot_dir,
    paste0("wri_score_", safe_name(region_name), ".png")
  )

  if (file.exists(out_path)) {
    return(out_path)
  }

  if (is.null(region$aoi)) {
    existing_full_plot <- file.path(demo_output_dir, "wri_score_full.png")

    if (file.exists(existing_full_plot)) {
      file.copy(existing_full_plot, out_path, overwrite = TRUE)
      return(out_path)
    }

    wri_score <- get_layer("WRI_score")
  } else {
    target_cache_path <- file.path(
      extract_cache_dir,
      safe_name(region_name),
      "WRI_score.tif"
    )

    existing_cache <- wri_extract_sizes[
      wri_extract_sizes$region == region_name &
        wri_extract_sizes$id == "WRI_score" &
        is.na(wri_extract_sizes$error) &
        !is.na(wri_extract_sizes$bytes),
      "file_path"
    ]
    existing_cache <- existing_cache[file.exists(existing_cache)]

    if (length(existing_cache) > 0L) {
      wri_score <- terra::rast(existing_cache[1])
    } else {
      wri_score <- get_layer(
        "WRI_score",
        aoi = region$aoi,
        aoi_crs = region$aoi_crs
      )

      dir.create(dirname(target_cache_path), recursive = TRUE, showWarnings = FALSE)
      terra::writeRaster(
        wri_score,
        target_cache_path,
        overwrite = TRUE,
        gdal = c("COMPRESS=DEFLATE", "TILED=YES", "BIGTIFF=IF_SAFER")
      )
    }
  }

  png(out_path, width = 1200, height = 900, res = 150)
  on.exit(dev.off(), add = TRUE)
  plot(
    wri_score,
    main = paste("WRI Score -", region_name),
    col = hcl.colors(100, "YlOrRd", rev = TRUE)
  )

  out_path
}

wri_layers <- wri_overview_df()
wri_layers <- subset(
  wri_layers,
  grepl("geotiff", asset_type, ignore.case = TRUE) &
    tolower(as.character(is_hosted)) == "true"
)
wri_layers <- wri_layers[!duplicated(wri_layers$id), ]

if (file.exists(detail_csv)) {
  wri_extract_sizes <- read.csv(detail_csv, stringsAsFactors = FALSE)
} else {
  wri_extract_sizes <- data.frame(
    region = character(),
    id = character(),
    file_path = character(),
    bytes = numeric(),
    error = character(),
    stringsAsFactors = FALSE
  )
}

for (region_name in names(wri_size_regions)) {
  region <- wri_size_regions[[region_name]]

  for (i in seq_len(nrow(wri_layers))) {
    layer_id <- wri_layers$id[i]
    matching_rows <- wri_extract_sizes$region == region_name &
      wri_extract_sizes$id == layer_id
    already_done <- matching_rows &
      is.na(wri_extract_sizes$error) &
      !is.na(wri_extract_sizes$bytes)

    if (any(already_done)) {
      next
    }

    message("Measuring ", region_name, " / ", layer_id)

    one_result <- tryCatch(
      measure_one_extract(
        layer_id = layer_id,
        href = wri_layers$asset_href[i],
        region_name = region_name,
        region = region
      ),
      error = function(e) {
        data.frame(
          region = region_name,
          id = layer_id,
          file_path = NA_character_,
          bytes = NA_real_,
          error = conditionMessage(e),
          stringsAsFactors = FALSE
        )
      }
    )

    wri_extract_sizes <- rbind(wri_extract_sizes, one_result)
    write.csv(wri_extract_sizes, detail_csv, row.names = FALSE)
  }
}

complete_sizes <- subset(wri_extract_sizes, is.na(error) & !is.na(bytes))
wri_extract_totals <- merge(
  data.frame(region = names(wri_size_regions), stringsAsFactors = FALSE),
  aggregate(bytes ~ region, complete_sizes, sum),
  by = "region",
  all.x = TRUE,
  sort = FALSE
)
wri_extract_totals$gigabytes <- wri_extract_totals$bytes / 1024^3

write.csv(wri_extract_totals, totals_csv, row.names = FALSE)

bar_values <- wri_extract_totals$gigabytes
bar_values_plot <- ifelse(is.na(bar_values), 0, bar_values)
bar_labels <- c(
  "Santa Barbara\nCounty",
  "Southern\nCalifornia",
  "California",
  "Full layer\nextent"
)

png(plot_path, width = 1200, height = 800, res = 150)
old_par <- par(mar = c(6.5, 5, 3, 1))
bars <- barplot(
  bar_values_plot,
  names.arg = bar_labels,
  ylab = "Total file size (GB)",
  main = "Total Size of All Hosted WRI Layers by Extract Region",
  col = c("#2C7FB8", "#41B6C4", "#A1DAB4", "#FEE08B"),
  border = NA,
  ylim = c(0, max(bar_values_plot, na.rm = TRUE) * 1.15),
  las = 1
)
text(
  x = bars,
  y = bar_values_plot,
  labels = ifelse(is.na(bar_values), "NA", format(round(bar_values, 2), nsmall = 2)),
  pos = 3,
  cex = 0.85
)
par(old_par)
dev.off()

wri_score_plot_paths <- vapply(
  names(wri_size_regions),
  function(region_name) {
    message("Saving WRI_score plot for ", region_name)

    tryCatch(
      save_wri_score_plot(region_name, wri_size_regions[[region_name]]),
      error = function(e) {
        warning(
          "Could not save WRI_score plot for ",
          region_name,
          ": ",
          conditionMessage(e),
          call. = FALSE
        )
        NA_character_
      }
    )
  },
  character(1)
)

wri_extract_totals
```

```{r wri-extract-size-plot, echo=FALSE, eval=file.exists("demo_output/wri_extract_size_totals.png") || file.exists("inst/demos/demo_output/wri_extract_size_totals.png")}
plot_path <- if (file.exists("demo_output/wri_extract_size_totals.png")) {
  "demo_output/wri_extract_size_totals.png"
} else {
  "inst/demos/demo_output/wri_extract_size_totals.png"
}
knitr::include_graphics(plot_path)
```

```{r wri-score-region-plots, echo=FALSE, eval=dir.exists("demo_output/wri_score_region_plots") || dir.exists("inst/demos/demo_output/wri_score_region_plots")}
wri_score_plot_dir <- if (dir.exists("demo_output/wri_score_region_plots")) {
  "demo_output/wri_score_region_plots"
} else {
  "inst/demos/demo_output/wri_score_region_plots"
}
wri_score_plot_files <- list.files(
  wri_score_plot_dir,
  pattern = "^wri_score_.*\\.png$",
  full.names = TRUE
)
knitr::include_graphics(wri_score_plot_files)
```

---

## Summary

We are working on avoiding wait times by cropping with SpatExtent exclusively, which terra treats as lazy windows over the remote COG, rather than SpatVectors, which materialize the cropped tiles.

### Functions

| Function | Purpose |
|----------|---------|
| `wri_overview()` | Print a summary of available layers |
| `wri_overview_df()` | Return available layers as a data frame |
| `layer_info(id)` | Full metadata for one layer |
| `query_stac_flexible(items, ...)` | Filter STAC items by property |
| `get_layer(id, aoi, aoi_crs)` | Stream a hosted raster, cropped to AOI |

### AOI input types for `get_layer()`

| AOI type | `aoi_crs` required? |
|----------|---------------------|
| `NULL` (returns full layer) | No |
| Numeric `c(xmin, ymin, xmax, ymax)` | Yes, always |
| File path to `.shp` | No (read from `.prj` sidecar) |
| File path to `.geojson` | No (embedded in file) |
| `sf` or `sfc` object | No (embedded in object) |
| `terra::SpatVector` | No (embedded in object) |
| `terra::SpatExtent` | Yes, always (terra strips the CRS) |
