---
title: "firex Package Demo"
format: html
---

# Getting Started

This demo shows how to explore, query, and retrieve WRI Aqueduct data using
the `firex` package. It covers all supported area-of-interest input types and
demonstrates validation, edge-case handling, and expected warning/error
messages.

## Data Sources

Spatial inputs used in this demo:

| File | Format | Source |
|------|--------|--------|
| `Eaton_Perimeter_20250121.shp` | Shapefile | [LA County Open Data — Eaton Fire Perimeter](https://data.lacounty.gov/datasets/lacounty::eaton-fire-perimeter/about) |
| `Eaton_Fire_Perimeter.geojson` | GeoJSON | [LA County Open Data — Eaton Fire Perimeter](https://data.lacounty.gov/datasets/lacounty::eaton-fire-perimeter/about) |

Both files are stored in `inst/demos/data/` and accessed via
`system.file("demos/data/...", package = "firex")`.

```{r setup, message=FALSE}
# Install from GitHub (once):
devtools::install_github("ignitR-package/firex")

# Development mode from the package root:
# devtools::load_all()

library(firex)
library(terra)
library(sf)

# Paths to bundled demo data (LA County Open Data, Eaton Fire perimeter)
data_dir  <- system.file("demos/data", package = "firex")
shp_path  <- file.path(data_dir, "Eaton_Perimeter_20250121.shp")
gjson_path <- file.path(data_dir, "Eaton_Fire_Perimeter.geojson")

# Pre-load as in-memory objects for the input-type examples below
eaton_sf   <- sf::st_read(shp_path,   quiet = TRUE)
eaton_vect <- terra::vect(shp_path)
eaton_ext  <- terra::ext(eaton_vect)   # SpatExtent (no embedded CRS)
```

---

## Step 1: Explore What's Available

`wri_overview()` reads the local STAC catalog and summarises available
domains, dimensions, and data types.

```{r overview}
catalog <- wri_overview()
catalog
```

`wri_overview_df()` returns a flat data frame for quick filtering.

```{r overview-df}
catalog_df <- wri_overview_df()
head(catalog_df[, c("id", "wri_domain", "wri_dimension", "data_type", "is_hosted")])
```

```{r explore-categories}
sort(unique(catalog_df$wri_domain))
sort(unique(catalog_df$wri_dimension))
sort(unique(catalog_df$data_type))
```

---

## Step 2: Filter and Inspect Layers

```{r filter-df}
# All water layers
subset(catalog_df, wri_domain == "water",
       select = c(id, wri_dimension, data_type))

# Indicator layers for air quality
subset(catalog_df, wri_domain == "air_quality" & data_type == "indicator",
       select = c(id, wri_dimension, data_type))

# Search by name pattern
catalog_df[grepl("surface_water|asthma", catalog_df$id),
           c("id", "wri_domain", "data_type")]
```

`layer_info()` returns full metadata for a single layer:

```{r layer-info}
layer_info("water_status_surface_water_quantity")
```

`query_stac_flexible()` filters the underlying STAC items by property:

```{r query-stac}
stac_items <- lapply(catalog$data$items, function(x) x$item)

water_items <- query_stac_flexible(stac_items,
                                   wri_domain = "water",
                                   data_type  = "indicator")
vapply(water_items, function(x) x$id, character(1))
```

---

## Step 3: Input Validation Helpers

Both `normalize_crs()` and `resolve_to_bbox()` return a length-1 logical
with attributes — they never throw. Inspect `isTRUE()` to branch on
success/failure, and `attr(result, "message")` on failure.

---

### `normalize_crs()` — Canonicalize a CRS specification

Accepts integer EPSG codes, `"EPSG:XXXX"` strings, bare numeric strings,
PROJ strings, and WKT.

```{r normalize-valid}
# Integer EPSG code
r_int <- normalize_crs(4326)
isTRUE(r_int);  attr(r_int, "crs")

# "EPSG:XXXX" string
r_str <- normalize_crs("EPSG:4326")
isTRUE(r_str);  attr(r_str, "crs")

# Bare numeric string (auto-prefixed)
r_bare <- normalize_crs("32611")
isTRUE(r_bare);  attr(r_bare, "crs")

# NULL is a valid no-op
r_null <- normalize_crs(NULL)
isTRUE(r_null)
```

```{r normalize-invalid}
# ── EDGE CASE: unrecognized CRS string ─────────────────────────────────────
r_bad <- normalize_crs("not-a-crs")
isTRUE(r_bad)
attr(r_bad, "message")
```

---

### `resolve_to_bbox()` — Extract a bbox from any spatial input

---

#### Input type: numeric vector `c(xmin, ymin, xmax, ymax)`

A CRS must always be supplied explicitly for numeric input.

```{r resolve-numeric}
res_num <- resolve_to_bbox(c(-118.2, 34.1, -117.9, 34.3),
                           aoi_crs = "EPSG:4326")
isTRUE(res_num)
attr(res_num, "bbox")
attr(res_num, "crs")
```

---

#### Input type: shapefile path (`.shp`)

CRS is read from the `.prj` sidecar; `aoi_crs` must be omitted.

```{r resolve-shp}
# Source: LA County Open Data — Eaton Fire Perimeter
res_shp <- resolve_to_bbox(shp_path)
isTRUE(res_shp)
attr(res_shp, "bbox")
attr(res_shp, "crs")
```

---

#### Input type: GeoJSON path (`.geojson`)

```{r resolve-geojson}
# Source: LA County Open Data — Eaton Fire Perimeter
res_gjson <- resolve_to_bbox(gjson_path)
isTRUE(res_gjson)
attr(res_gjson, "bbox")
attr(res_gjson, "crs")
```

---

#### Input type: `sf` object

```{r resolve-sf}
# eaton_sf loaded above via sf::st_read(); CRS is embedded — omit aoi_crs
res_sf <- resolve_to_bbox(eaton_sf)
isTRUE(res_sf)
attr(res_sf, "bbox")
```

---

#### Input type: `terra::SpatVector`

```{r resolve-spatvector}
# eaton_vect loaded above via terra::vect(); CRS is embedded — omit aoi_crs
res_vect <- resolve_to_bbox(eaton_vect)
isTRUE(res_vect)
attr(res_vect, "bbox")
```

---

#### Input type: `terra::SpatExtent`

`SpatExtent` objects carry no CRS, so `aoi_crs` is always required.

```{r resolve-spatextent}
# eaton_ext derived from eaton_vect above (terra::ext strips the CRS)
res_ext <- resolve_to_bbox(eaton_ext, aoi_crs = "EPSG:4326")
isTRUE(res_ext)
attr(res_ext, "bbox")
```

---

## Step 4: Retrieve and Plot Data with `get_layer()`

`get_layer(id, aoi, aoi_crs)` accepts the same AOI inputs as
`resolve_to_bbox()`. All reprojection from the AOI CRS to the raster's
native CRS is handled internally.

---

#### Input type: `aoi = NULL` — full global layer

```{r get-full}
t_full <- system.time(wri_full <- get_layer("WRI_score"))
t_full
# Plotting the full extent is slow; left commented out.
# plot(wri_full, main = "WRI Score — Full Extent",
#      col = hcl.colors(100, "YlOrRd", rev = TRUE))
```

---

#### Input type: numeric bbox + explicit CRS

```{r get-bbox}
bbox_norcal <- c(-122, 37, -121, 38)

t_num <- system.time(
  wri_norcal <- get_layer("WRI_score",
                           aoi     = bbox_norcal,
                           aoi_crs = "EPSG:4326")
)
t_num
```

```{r plot-bbox}
plot(wri_norcal,
     main = "WRI Score — Northern California (numeric bbox)",
     col  = hcl.colors(100, "YlOrRd", rev = TRUE))
```

---

#### Input type: shapefile path

```{r get-shp}
# Source: LA County Open Data — Eaton Fire Perimeter
t_shp <- system.time(
  wri_shp <- get_layer("WRI_score", aoi = shp_path)
)
t_shp
```

```{r plot-shp}
plot(wri_shp,
     main = "WRI Score — Eaton Fire Area (shapefile path)",
     col  = hcl.colors(100, "YlOrRd", rev = TRUE))
```

---

#### Input type: GeoJSON path

```{r get-geojson}
# Source: LA County Open Data — Eaton Fire Perimeter
t_gjson <- system.time(
  wri_gjson <- get_layer("WRI_score", aoi = gjson_path)
)
t_gjson
```

```{r plot-geojson}
plot(wri_gjson,
     main = "WRI Score — Eaton Fire Area (GeoJSON path)",
     col  = hcl.colors(100, "YlOrRd", rev = TRUE))
```

---

#### Input type: `sf` object

```{r get-sf}
t_sf <- system.time(
  wri_sf <- get_layer("WRI_score", aoi = eaton_sf)
)
t_sf
```

---

#### Input type: `terra::SpatVector`

```{r get-spatvector}
indicator_id <- "water_status_surface_water_quantity"

t_vect <- system.time(
  water_qty <- get_layer(indicator_id, aoi = eaton_vect)
)
t_vect
```

```{r plot-spatvector}
plot(water_qty,
     main = "Surface Water Quantity — Eaton Fire Area (SpatVector)",
     col  = hcl.colors(100, "YlOrRd", rev = TRUE))
```

---

#### Input type: `terra::SpatExtent`

```{r get-spatextent}
t_ext <- system.time(
  wri_ext <- get_layer("WRI_score",
                        aoi     = eaton_ext,
                        aoi_crs = "EPSG:4326")
)
t_ext
```

---

## Step 5: Edge Cases — Validation Failures, Warnings, and Errors

These examples demonstrate the guard-rail behavior of each function. None of
the cells below crash the document: validation-failure cells return `FALSE`
with an informative message, and error-triggering calls are wrapped in
`tryCatch`.

---

### Validation failures — `normalize_crs()`

```{r edge-normalize}
# ── Unrecognized CRS ───────────────────────────────────────────────────────
r <- normalize_crs("EPSG:999999999")   # valid format, invalid code
isTRUE(r);  attr(r, "message")
```

---

### Validation failures — `resolve_to_bbox()`

```{r edge-resolve-no-crs}
# ── Numeric bbox without aoi_crs ───────────────────────────────────────────
# get_layer would forward the same error here; resolve_to_bbox catches it early
r <- resolve_to_bbox(c(-118.2, 34.1, -117.9, 34.3))
isTRUE(r);  attr(r, "message")
```

```{r edge-resolve-crs-conflict}
# ── aoi_crs supplied when the object already carries a CRS ─────────────────
# sf objects embed their CRS; passing aoi_crs too is ambiguous and rejected
r <- resolve_to_bbox(eaton_sf, aoi_crs = "EPSG:4326")
isTRUE(r);  attr(r, "message")
```

```{r edge-resolve-extent-no-crs}
# ── SpatExtent without aoi_crs ─────────────────────────────────────────────
# SpatExtent strips the CRS; aoi_crs is mandatory
r <- resolve_to_bbox(eaton_ext)
isTRUE(r);  attr(r, "message")
```

---

### Error — invalid layer id

```{r edge-bad-id}
# ── Layer id not in catalog → stop() ───────────────────────────────────────
tryCatch(
  get_layer("this_layer_does_not_exist"),
  error = function(e) cat("Error:", conditionMessage(e), "\n")
)
```

---

### Warning — bbox partially overlaps the layer extent

When the requested bbox extends beyond the layer's footprint, `get_layer`
emits a warning describing which sides overflow and then crops to the
available intersection. The result is still returned.

```{r edge-partial-overlap, warning=TRUE}
# ── Bbox that extends beyond the layer's western/eastern/northern boundary ──
# get_layer emits a warning describing which sides overflow, then crops to
# the available intersection and returns the raster normally.
bbox_overflow <- c(-130, 30, -55, 55)   # well beyond the contiguous US

wri_overflow <- get_layer("WRI_score",
                           aoi     = bbox_overflow,
                           aoi_crs = "EPSG:4326")
```

---

### Error — bbox has no overlap with the layer extent

```{r edge-no-overlap}
# ── Bbox entirely outside the layer → stop() ───────────────────────────────
# Antarctica is far outside any US-only or regional WRI layer
bbox_antarctica <- c(-80, -90, -60, -70)

tryCatch(
  get_layer("WRI_score", aoi = bbox_antarctica, aoi_crs = "EPSG:4326"),
  error = function(e) cat("Error:", conditionMessage(e), "\n")
)
```

---

## Summary

| Step | Function(s) | Purpose |
|------|------------|---------|
| 1 | `wri_overview()`, `wri_overview_df()` | Browse the catalog |
| 2 | `layer_info()`, `query_stac_flexible()` | Inspect layer metadata |
| 3 | `normalize_crs()`, `resolve_to_bbox()` | Validate inputs before fetching |
| 4 | `get_layer(id, aoi, aoi_crs)` | Stream a hosted raster, optionally cropped |
| 5 | — | Edge cases: validation failures, partial-overlap warnings, no-overlap errors |

**`aoi` input types accepted by `resolve_to_bbox()` and `get_layer()`:**

| Type | `aoi_crs` required? |
|------|---------------------|
| `NULL` (full layer) | — |
| Numeric `c(xmin, ymin, xmax, ymax)` | Yes |
| Shapefile path (`.shp`) | No (read from `.prj`) |
| GeoJSON path (`.geojson`) | No (embedded in file) |
| `sf` / `sfc` object | No (embedded in object) |
| `terra::SpatVector` | No (embedded in object) |
| `terra::SpatExtent` | Yes (CRS is stripped by terra) |
